\documentclass{cit_thesis}
\usepackage{amsmath, amssymb, braket, units,bm,bbm}
\usepackage[mathscr]{euscript}
\usepackage{graphicx}
\usepackage{subfigure}		% This is supposed to be obsolete, superseded by subfig
\usepackage{rotating, multirow, mathtools}
\usepackage{epstopdf, times}

\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,breaklinks]{hyperref}
\usepackage{breakurl}
\usepackage{color}
\usepackage{wrapfig}

%macros for intro
\newcommand{\lct}{\epsilon^{\mu\nu\lambda}}
\newcommand{\lcf}{\epsilon^{\mu\nu\lambda\sigma}}
%macros for FQHE
\newcommand{\JJ}{\vec{J}}
\newcommand{\QQ}{\vec{Q}}
\def\cJ{J}
\def\cQ{Q}
\def\cF{F}
\def\cG{ G}
\newcommand{\FF}{\vec{F}}
\newcommand{\GG}{\vec{G}}
\newcommand{\genC}{C^{\mu\nu}_{ab}(k)}
\def\vcJ{{\vec\cJ}}
\def\vcQ{{\vec\cQ}}
\def\vcF{{\vec\cF}}
\def\vcG{{\vec\cG}}

\def\br{{\bf r}}
\def\bR{{\bf R}}
\def\bx{{\bf x}}
\def\by{{\bf y}}
\def\bj{{\bf j}}
\def\bk{{\bf k}}
\def\valpha{{\vec{\alpha}}}
\def\vbeta{{\vec{\beta}}}
\def\vgamma{{\vec{\gamma}}}
\def\up{\uparrow}
\def\dn{\downarrow}


\def\half{{\frac{1}{2}}}
\def\Aext{A^{\rm ext}}
\def\vAext{{\vec{A}^{\rm ext}}}

\newcommand{\curl}{\vec{\nabla}\times}
\newcommand{\divv}{\vec{\nabla}\cdot}
\newcommand{\uxu}{U(1)\times U(1)}
\def\ra{\rangle} % bra
\def\la{\langle} % ket

%macros for SO34D
\newcommand{\scripty}[1]{w}
\def\uz{U(1)\times\mathbb{Z}_2}
\def\ztwo{\mathbb{Z}_2}
\def\ztwot{\mathbb{Z}_2^T}
\newcommand{\cp}{$CP^1$ }
\newcommand{\nccp}{$NCCP^1$ }

\def\cP{{P}}
\def\hext{h^{\rm ext}}
\def\uu{B}
\def\MM{N}
\def\mm{n}

%macros for bilayer
\newcommand{\corrR}{{g}}
\newcommand{\corrM}{{\tilde g}}
\renewcommand{\r}{{\mathbf{r}}}
\newcommand{\shift}{\mathscr{S}}
\newcommand{\D}{\mathrm{d}}
\newcommand{\DSAS}{\Delta_\textrm{SAS}}
\newcommand{\Dnorm}{\DSAS/\frac{e^2}{\epsilon\ell_B}}		% change to (\frac{e^2}{\epsilon\ell_B})
\newcommand{\barthird}{\ensuremath{\overline{1/3}}}
\newcommand{\Uone}{\ensuremath{\mathrm{U(1)}} }
\newcommand{\SU}[1]{\ensuremath{\mathrm{SU(#1)}}}
\newcommand{\spinon}{{\phi_s}}

\begin{document}
\title{Numerical Studies of Topological phases}
\author{Scott Geraedts}


\begin{abstract}
The topological phases of matter have been a major part of condensed matter physics research since the discovery of the quantum Hall effect in the 1980s. Historically, much of this research has focused on the study of systems of free fermions, such as the integer quantum Hall effect, quantum spin Hall effect, and topological insulator. Though these free fermion systems can play host to a variety of interesting phenomena, the physics of interacting topological phases is even richer. Unfortunately, there is a shortage of theoretical tools that can be used to approach interacting problems. In this thesis I will discuss progress in using two different numerical techniques to study topological phases.

Recently much research in topological phases has focused on phases made up of bosons. Unlike fermions, free bosons form a condensate and so interactions are vital if the bosons are to realize a topological phase. Since these phases are difficult to study, much of our understanding comes from exactly solvable models, such as Kitaev's toric code, as well as Levin-Wen and Walker-Wang models. We may want to study systems for which such exactly solvable models are not available. In this thesis I present a series of models which are not solvable exactly, but which can be studied in sign-free Monte Carlo simulations. The models work by binding charges to point topological defects. They can be used to realize bosonic interacting versions of the quantum Hall effect in 2D and topological insulator in 3D. Effective field theories of ``integer'' (non-fractionalized) versions of these phases were available in the literature, but our models also allow for the construction of fractional phases. We can measure a number of properties of the bulk and surface of these phases. 

Few interacting topological phases have been realized experimentally, but there is one very important exception: the fractional quantum Hall effect (FQHE). Though the fractional quantum Hall effect we discovered over 30 years ago, it can still produce novel phenomena. Of much recent interest is the existence of non-Abelian anyons in FQHE systems. Though it is possible to construct wave functions that realize such particles, whether these wavefunctions are the ground state is a difficult quantitative question that must be answered numerically. In this thesis I describe progress using a density-matrix renormalization group algorithm to study a bilayer system thought to host non-Abelian anyons. We find phase diagrams in terms of experimentally relevant parameters, and also find evidence for a non-Abelian phase known as the `interlayer Pfaffian'.
%A diverse set of topological phases are studied using a variety of numerical techniques. First, a formalism for studying a class of bosonic models, both analytically and in Monte Carlo simulations, is developed. This formalism is applied to realize bosonic versions of the integer and fractional bosonic quantum Hall effects. This phase is realized by binding charges to point topological defects in two dimensions (vortices). Properties of this phase, such as the Hall conductivity, charge and statistics of quasiparticles, and gaplessness of the edge modes, are determined. The idea of binding charges to point topological defects is then extended to three dimensions, where a bosonic topological insulating phase is realized, along with its fractional version. The identity of this topological phase is determined by observing a `Witten effect', as well as by studying its surface physics. Lastly, infinite DMRG methods are applied to the study of a bilayer fractional quantum Hall system. A number of different phases are realized, including the non-Abelian `interlayer Pfaffian' phase. These phases are identified by studying the charges of the elementary quasiparticles, as well as the entanglement spectra and momentum polarization.
\end{abstract}

\maketitle
\tableofcontents

\mainmatter
\include{intro}
\include{formalism}
\include{FQHE}
\include{SO34D}
\include{bilayer}
\include{conclud}

\bibliographystyle{ieeetr}
\bibliography{thesis}
\end{document}

convert X.png eps3:X.eps
lpr -o Duplex=DuplexNoTumble -o page-ranges=7-21 master.pdf
